---
title: "bruneians by age and resident status"
description: "hellow"
author: "Matthew Ng"
type: default
date: "2024-05-14"
categories: [brunei, statistics]
format:
  html:
    code-fold: true
    code-summary: "show code"
---

### libraries

Main libraries in code block.

```{r library, echo = T, message = F, warning = F}
library(librarian)
shelf(readxl, tidyverse, reshape2)
```

# introduction

Last week, I found myself back in Brunei—not for the usual family visit, but for a series of work meetings and to give a talk on data science.

It was a blink-and-you-miss-it trip; barely three days on the ground.

In true Bruneian fashion, we spent the time between meetings and talks dining out. By our fourth restaurant visit, my colleague Thomas—a Kiwi and a first-timer in Brunei—asked in the most hilariously earnest way: "Why aren't there any locals running their own businesses?"

To be honest, I didn't know. I spent most of my first 18 years there, and it had always been like this. I knew we had a population of around half a million, but that was about all I could offer him.

# data

That was the inspiration for this post.

I managed to collate and convert the terrible .pdf from the Department of Statistics into some usable tables. If anyone stumbles upon this blog, they might want to recreate this.

The main data sets I compiled include the age group by nationality table (BN_age_group_nationality.xlsx in this post's repo) from the published 2021 national census.

The main data structure is pretty simple:

1.  `age_group`: Mostly in 5-year divisions, top-coded at 85+

2.  `type`: Categorical options include total, Bruneian, permanent_resident, and temporary_resident

3.  `male`: Total number of males

4.  `female`: Total number of females

```{r}
# main data structure
read_xlsx("BN_age_group_nationality.xlsx") %>% 
  str()
```

## processing

The only real processing we need to do is to get the lower age limit for each of the age groups within the data set. That's relatively simple with the following code:

```{r}
# read as age_group object
age_group <- read_xlsx("BN_age_group_nationality.xlsx")

# get first two characters from column
age_group <- age_group %>%
  mutate(lower_age = substr(age_group, 1, 2)) %>% 
  # reposition to second column
  relocate(lower_age, .before = age_group)

# final dataset
age_group
```

# pyramids

To get a better sense of our population in 2021, we'll create three charts: a population pyramid by age group, a faceted pyramid by nationality, and a gender group per capita by nationality graph.

## population by age group

In the grand scheme of things, men outnumber women in Brunei by around 11.4%. What's particularly interesting is the noticeable bulge in the 20-45 age range for men. In this subset, men outnumber women by an average of about 21.5%, compared to the overall difference per age group of around 4.6%.

```{r message = F, warning = F, echo = T}
# total population by age group
age_group %>%
  filter(type == "total") %>% 
  # melt for male and female
  melt(value.name = "total",
       variable.name = "gender") %>%
  # plot
  ggplot(aes(fill = gender, 
             x = lower_age, 
             y = ifelse(
               test = gender == "male",
               # flip the values to the negative
               yes = -total,
               no = total),
             )) +
         geom_col(alpha = 0.9) +
  # flip the y-axis
  coord_flip() +
  scale_y_continuous(
    labels = abs, # set as absolute to remove negative sign
    limits = c(-25000, 25000) # set limits aesthetics
    ) +
  labs(
    x = "age group",
    y = "total",
    title = "population by age group"
  ) + # change color with brewer
  scale_fill_brewer(palette = "Set1") +
  theme_minimal()
```

## age groups by nationality

This is where the real issue becomes clear. Our local population is relatively young, and it seems we are compensating for the shortage of young Bruneians with temporary residents.

```{r message = F, warning = F, echo = T}
# total population by age group
age_group %>%
  #filter out total
  filter(type != "total") %>%
  # melt for male and female
  melt(
    value.name = "total",
    variable.name = "gender") %>%
  # plot
  ggplot(aes(fill = gender, 
             x = lower_age, 
             y = ifelse(
               test = gender == "male",
               # flip the values to the negative
               yes = -total,
               no = total),
             )) +
         geom_col(alpha = 0.9) +
  # flip the y-axis
  coord_flip() +
  scale_y_continuous(
    labels = abs, # set as absolute to remove negative sign
    # limits = c(-30000, 30000) # set limits aesthetics
    ) +
  labs(
    x = "age group",
    y = "total",
    title = "population by nationality"
  ) + # change color with brewer
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(~type) +
  theme_minimal()
```

Temporary residents make up a significant proportion of the general population, constituting up to 34.4% across both genders. Among men aged 20 to 49, this percentage can reach as high as 40% in certain age groups.

```{r message = F, warning = F, echo = T}
age_group %>% 
  filter(type != "total") %>%
  melt(value.name = "total",
       variable.name = "gender") %>% 
  # filter(age_group %in% c("20-24", "25-29", "30-34", "35-39", "40-44")) %>% 
  group_by(age_group, type) %>%
  summarise(total = sum(total)) %>% 
  # pivot wider
  pivot_wider(names_from = type, values_from = total) %>% 
  mutate(total = bruneian + permanent_resident + temporary_resident) %>% 
  # percentage of total
  mutate(
    bruneian = bruneian/total * 100,
    permanent_resident = permanent_resident/total * 100,
    temporary_resident = temporary_resident/total * 100) %>% 
  select(-total)
```

Among men aged 20 to 49, this percentage can reach as high as 40% in certain age groups.

```{r message = F, warning = F, echo = T}
age_group %>% 
  filter(type != "total") %>%
  melt(value.name = "total",
       variable.name = "gender") %>% 
  filter(age_group %in% c("20-24", 
                          "25-29", 
                          "30-34", 
                          "35-39", 
                          "40-44", 
                          "45-49")) %>% 
  group_by(age_group, type, gender) %>%
  filter(gender == "male") %>% 
  summarise(total = sum(total)) %>% 
  # pivot wider
  pivot_wider(names_from = type, values_from = total) %>% 
  mutate(total = bruneian + permanent_resident + temporary_resident) %>% 
  # percentage of total
  mutate(
    bruneian = bruneian/total * 100,
    permanent_resident = permanent_resident/total * 100,
    temporary_resident = temporary_resident/total * 100) %>% 
  select(-total)
```

## per capita of gender and nationality

The proportion of temporary residents in the 20-45 age group is pretty huge, with over 70% of them falling into this range. 

```{r message = F, warning = F}
shelf(gganimate, png, gifski)

pop_pyr <- age_group %>% 
  filter(type != "total") %>% 
  group_by(type) %>%
  mutate(total_m = sum(male),
         total_f = sum(female),
         `per capita m` = (male/total_m) *  100,
         `per capita f` = (female/total_f) *  100) %>%
  select(-male, -female, -total_m, -total_f) %>% 
  melt(variable.name = "gender",
       value.name = "total") %>% 
  ggplot(aes(fill = gender, 
             x = lower_age, 
             y = ifelse(
               test = gender == "per capita m",
               # flip the values to the negative
               yes = -total,
               no = total),
             )) +
  geom_col(alpha = 0.9) +
  coord_flip() +
  scale_y_continuous(
    labels = abs, # set as absolute to remove negative sign
    limits = c(-18, 18) # set limits aesthetics
    ) +
  labs(
    x = "age group",
    y = "% per capita",
    title = "% per capita by age group/nationality"
  )  + 
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() + # animate
  transition_states(type, transition_length = 2, state_length = 1)
  
pop_pyr
```

This big chunk of the population raises some interesting questions about what their presence means for us. The main question is whether we should let these temporary residents stay as permanent residents or even citizens to keep our population and workforce stable. If we need these workers long-term to support the economy, what are the effects of integrating them into our society? Are we ready to support their integration, and will they fit into the community in a way that feels right to Bruneians?

Also, since many of these temporary residents spend most of their prime working years in Brunei, should there be a way for them to become citizens since they've contributed to nation-building? How will our migration policies change, especially as our mostly young Bruneian population gets older and starts filling these age groups/labour roles?

Going back to the original question: why aren't there more locals running their own businesses? Maybe Thomas was only seeing part of the picture. It’s tough to say since there are many factors at play. It seems to be a mix of having a young local population and a cultural aversion for front-facing or physical labor roles. It would be interesting to find out which parts of the workforce these temporary residents are in—especially the men. What industries are they in, and are they taking on roles that the local (predominantly male) workforce can't handle?