---
title: "bruneians by age and resident status"
description: "hellow"
author: "Matthew Ng"
type: default
date: "2024-05-14"
categories: [brunei, statistics]
format:
  html:
    code-fold: true
    code-summary: "show code"
---

### libraries

Main libraries in code block.

```{r library, echo = T, message = F, warning = F}
library(librarian)
shelf(readxl, tidyverse, reshape2)
```

# introduction

Last week, I found myself back in Brunei—not for the usual family visit, but for a bunch of work meetings and to give a talk on data science.

It was a real blink-and-you-miss-it trip; barely three days on the ground.

In true Bruneian fashion, we filled the time between meetings and talks by dining out. I think it was by our fourth restaurant visit that my colleague Thomas—a Kiwi and a first-timer in Brunei—asked in the most hilariously earnest way: "why aren't there any locals running their own businesses?"

To be honest, I didn't know --- I spent most of my first 18 years there, and it had always really been this . I mean  knew we had a population of more or less half a million, but that was really the limit of what I could offer him.

# data

That was the inspiration for this post.

I managed to collate and convert the terrible `.pdf` from the Department of Statistics into some usable tables.

Thinking if anyone stumbled upon this blog, they'd maybe want to recreate this.

The main data sets I collated from the table is the age group by nationality table (`BN_age_group_nationality.xlsx` in this post's repo) from the published 2021 national census.

The main data structure is pretty simple:

1.  `age_group`: Mostly at 5-year division, top-coded at 85+
2.  `type`: categorical options at total, bruneian, permanent_resident, and temporary_resident
3.  `male`: total number of males
4.  `female`: total number of females

```{r}
# main data structure
read_xlsx("BN_age_group_nationality.xlsx") %>% 
  str()
```

## processing

The only real processing we need to do is to get the lower age limit for each of the age groups within the data set. That's relatively simple with the following code:

```{r}
# read as age_group object
age_group <- read_xlsx("BN_age_group_nationality.xlsx")

# get first two characters from column
age_group <- age_group %>%
  mutate(lower_age = substr(age_group, 1, 2)) %>% 
  # reposition to second column
  relocate(lower_age, .before = age_group)

# final dataset
age_group
```

# pyramids

To get a better sense of our population in 2021, we'll make 3 pyramids for the total by age group, a facet by nationality, and a per capita by nationality graph.

## population by age group

In the grand scheme of things, men outnumber women in Brunei by around 11.4%. What I find particularly interesting is the odd bulge at around 20-45 for men. At this subset, men outnumbered women by an average of about 21.5%, compared to the overall difference per age group of around 4.6%.

```{r message = F, warning = F, echo = T}
# total population by age group
age_group %>%
  filter(type == "total") %>% 
  # melt for male and female
  melt(value.name = "total",
       variable.name = "gender") %>%
  # plot
  ggplot(aes(fill = gender, 
             x = lower_age, 
             y = ifelse(
               test = gender == "male",
               # flip the values to the negative
               yes = -total,
               no = total),
             )) +
         geom_col(alpha = 0.9) +
  # flip the y-axis
  coord_flip() +
  scale_y_continuous(
    labels = abs, # set as absolute to remove negative sign
    limits = c(-25000, 25000) # set limits aesthetics
    ) +
  labs(
    x = "total",
    y = "age group",
    title = "population by age group"
  ) + # change color with brewer
  scale_fill_brewer(palette = "Set1") +
  theme_minimal()
```

## age groups by nationality

This is where we really see the reason why.

```{r message = F, warning = F, echo = T}
# total population by age group
age_group %>%
  #filter out total
  filter(type != "total") %>%
  # melt for male and female
  melt(
    value.name = "total",
    variable.name = "gender") %>%
  # plot
  ggplot(aes(fill = gender, 
             x = lower_age, 
             y = ifelse(
               test = gender == "male",
               # flip the values to the negative
               yes = -total,
               no = total),
             )) +
         geom_col(alpha = 0.9) +
  # flip the y-axis
  coord_flip() +
  scale_y_continuous(
    labels = abs, # set as absolute to remove negative sign
    # limits = c(-30000, 30000) # set limits aesthetics
    ) +
  labs(
    x = "total",
    y = "age group",
    title = "population by age group"
  ) + # change color with brewer
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(~type) +
  theme_minimal()
```

We can see here that temporary residents make up quite a large proportion of the general population. In fact, overall, they constitute up to 34.4% across both genders. 

```{r message = F, warning = F, echo = T}
age_group %>% 
  filter(type != "total") %>%
  melt(value.name = "total",
       variable.name = "gender") %>% 
  # filter(age_group %in% c("20-24", "25-29", "30-34", "35-39", "40-44")) %>% 
  group_by(age_group, type) %>%
  summarise(total = sum(total)) %>% 
  # pivot wider
  pivot_wider(names_from = type, values_from = total) %>% 
  mutate(total = bruneian + permanent_resident + temporary_resident) %>% 
  # percentage of total
  mutate(
    bruneian = bruneian/total * 100,
    permanent_resident = permanent_resident/total * 100,
    temporary_resident = temporary_resident/total * 100) %>% 
  select(-total)
```
Among men aged 20 to 49, the percentage of temporary residents can reach as high as 40% in certain age groups.

```{r message = F, warning = F, echo = T}
age_group %>% 
  filter(type != "total") %>%
  melt(value.name = "total",
       variable.name = "gender") %>% 
  filter(age_group %in% c("20-24", 
                          "25-29", 
                          "30-34", 
                          "35-39", 
                          "40-44", 
                          "45-49")) %>% 
  group_by(age_group, type, gender) %>%
  filter(gender == "male") %>% 
  summarise(total = sum(total)) %>% 
  # pivot wider
  pivot_wider(names_from = type, values_from = total) %>% 
  mutate(total = bruneian + permanent_resident + temporary_resident) %>% 
  # percentage of total
  mutate(
    bruneian = bruneian/total * 100,
    permanent_resident = permanent_resident/total * 100,
    temporary_resident = temporary_resident/total * 100) %>% 
  select(-total)
```
## per capita by nationality
Final pyramids show per capita of the age group and gender. It is the most telling of the nation's overall population structure.

We can draw our own conclusions here on the implications on the Bruneian system.
```{r message = F, warning = F}
age_group %>% 
  filter(type != "total") %>% 
  group_by(type) %>% 
  mutate(total_m = sum(male),
         total_f = sum(female),
         `per capita male` = (male/total_m) * 100,
         `per capita female` = (female/total_f) * 100) %>% 
  select(lower_age, `per capita male`, `per capita female`) %>% 
  melt(variable.name = "gender",
       value.name = "total") %>% 
  ggplot(aes(fill = gender, 
             x = lower_age, 
             y = ifelse(
               test = gender == "per capita male",
               # flip the values to the negative
               yes = -total,
               no = total),
             )) +
         geom_col(alpha = 0.9) +
  # flip the y-axis
  coord_flip() +
  facet_wrap(~type) +
  scale_y_continuous(
    labels = abs, # set as absolute to remove negative sign
    limits = c(-20, 20) # set limits aesthetics
    ) +
  labs(
    x = "total",
    y = "age group",
    title = "population by age group"
  ) + # change color with brewer
  scale_fill_brewer(palette = "Set1") +
  theme_minimal()
```
